---
title: "Creating and Working with Basic R Data Structures: Vector & Data Frame"
output: html_notebook
---
## **Activity 1**
## Vectors

### Create variables
Store the vector `1, 2, 3, 4, 5, 4, 3, 2, 0` as variable `x`

```{r}

```

View the vector
```{r}

```

### Compare values in a vector
What is the value of `x > 3`
```{r}

```


### Vector operations
Create and store these vectors:

`seq(1, 10 by = 2)` as variable `a`

`10` as variable `b`

`10, 100, 1000` as variable `c`

Vector `a`
```{r}

```


Vector `b`
**TIP: ( b <- object ) make <- not print invisibly so you can see the value being assigned automatically without needing to type the object name after assigning it**
```{r}

```


Vector `c`
```{r}

```

What is the value of `a + b`?
```{r}

```

What is the value of `a + c`?
```{r}

```

## Object class and other properties
Create and store three vectors:

`'Brandon', 'Carla', 'Kevin', 'Taylor', 'Steven', 'Riley'` as `name`

`22, 25, 18, 38, 12, 20` as `age`

`5.5, 6.1, 4.9, 6.6, 4.3, 5.5` as `height`

`'M', 'F', 'M', 'X', 'M', 'X'` as `gender`

```{r}

```


### Class
What is the object class for `name` and `age`?
```{r}

```


### Structure
Display the internal structure for `height` and `gender`?
```{r}

```


### Dimension
use `length()` to check the length of the `name` vector.
```{r}
length(name)
```



### Check data types
What is the value of `is.numeric(name)`?
```{r}
is.numeric(name)
```


What is the value of `is.character(name)`?
```{r}
is.character(name)
```


What is the value of `is.vector(name)`?
```{r}
is.vector(name)
```


What is the value of `is.data.frame(name)`?
```{r}
is.data.frame(name)
```



### Explicit and implicit object conversion

Use 'as.character()` to convert the numeric vector `age` to a character vector explicitly and store it as `age.char`
```{r}

```



How many friends are older than 21?

Using the numeric `age` vector
```{r}

```

Using the character `age.char` vector
```{r}

```

Using a non-sensical vector for comparison such as `name`
```{r}

```





 
## **Activity 2**
### Accessing the elements in a vector

What is the name of the third person included in the `name` vector?
Hint: Use `name[<insert-index-here>]`
```{r}

```

What are the names of the 2nd, 3rd, 4th, and 5th person on the names vector?

Use the `name` vector and specify which index to KEEP so that only the 2nd, 3rd, 4th, and 5th person are printed
```{r}

```

Use the `name` vector and specify which index to EXCLUDE so that only the 2nd, 3rd, 4th, and 5th person are printed
```{r}

```


## **Data Frames**

### Create a data frame from vectors
Create a data frame from the `name, age, height, gender` vectors as `friends`
```{r}

```


## Cleaning the workspace
The `name, age, height, gender` vectors are redundant now that there  is a `friends` data frame. Remove these vectors from the workspace with `rm(<object>)`.


```{r}

```


Check what other objects are in the workspace with `ls()`. 
```{r}

```

Moving forward you only need the friends data frame. Remove all other objects. 
```{r}

```


### Overview of data frame

#### Use base R functions
Use `summary()` to learn more about the data frame
```{r}

```


Use `str()` to learn more about the data frame
```{r}


```

View the variables in a data frame with `names()`
```{r}

```


### View observations 
View the first observation in the data frame with `head(<data frame>, 1)`
```{r}

```

View the last observation in the data frame with `tail(<data frame>, 1)`
```{r}

```

## Object class and other properties

### Dimension

How does value of `length(friends)` compared to  `dim(friends)` ?

`length()` value for friends data frame
```{r}


```

`dim()` value for friends data frame
```{r}


```

What is the value of `nrow(friends)`?
```{r}


```

What is the value of `ncol(friends)`?
```{r}

```


### Check data types
What is the value of `is.numeric(friends)`?
```{r}

```

What is the value of `is.character(friends)`?
```{r}

```

What is the value of `is.vector(friends)`?
```{r}

```

What is the value of `is.data.frame(friends)`?
```{r}

```



## Accessing the vectors in a data frame

How many friends are older than 21?
```{r}

```

Use the logical operators along with `table()` to count how many people are older than 21
```{r}

```


Create a crosstab of friends older than 21 by gender using `table(<gender vector>,<logical vector for age older than 21>)`
```{r}


```



What output is produced by `plot(<age>)
```{r}

```



What output is produced by `plot(<gender vector>)`?
```{r}

```


What output is produced by `plot(<name vector>)`?
```{r}

```


What is the output of `plot(<age vector>,<height vector>)?
```{r}

```

Create a boxplot of height with `boxplot()` to view the distribution
```{r}

```




## **Activity 3**

#### Use package functions
Use the function `glimpse()` provided by the `dplyr` package to learn more about the data frame friends.

Install the dplyr package if it's not already installed (`install.packages("dplyr")`) and load the package `library("dplyr")`. 

```{r}

```



When a package is loaded all of its functions are loaded to the current memory. Suppose you want to use  `describe()` once and otherwise do not need other functions from the Hmisc package. One alternative is to access that function without loading the package. That can be done by specifying `packagename::functionname()`. (Note: Package installation is still needed)

Use the `describe()`function without loading the `Hmisc` package
```{r}

```

> **How does the overview of the friends data frame differ when using different functions: summary(), str(), glimpse(),describe()?**

> **How does the output differ for character and numeric vectors?**

> **Why might we want to refer to a package function explicity (e.g.`Hmisc::describe()` versus implicitly (e.g. loading with `library("Hmisc")` and then calling function `describe())?**

Hint: What messages did R print in the console after loading the package `dplyr`? What happens if you load the Hmisc package now after having loaded the dplyr package earlier?


### A different way to clean the workspace - Demo
Earlier we used `ls()` to check the objects in the workspace and used `rm()` to remove any object in the workspace that was not our `friends` data frame. 

Recreate the objects since we removed them earlier
```{r}
a <- seq(1, 10, by = 2)
b <- c(10)
c <- c(10, 100, 1000)
x <-  c(1, 2, 3, 4, 5, 4, 3, 2, 0)
```

Use `ls()` to view the objects in the workspace
```{r}
ls()
```

What is the class of the value generated by `ls()`
```{r}
class(ls())
```

Use an index to select only the `"friends"` value from the output returned by ls()
```{r}
ls()[4]
```

Return a logical vector indicating TRUE for any element from `ls()` that is not "friends" and FALSE if it is friends. Store the logical vector as `isnot.friends`.
```{r}
isnot.friends <- NULL # Place holder. Add name to workspace without any value.
( isnot.friends <- ls() != "friends" ) 
```


Use `ls()` and `isnot.friends` to select the elements from `ls()` that are not friends. We want a character vector that excludes the 'friends' value. 
```{r}
ls()[isnot.friends]
```

Use `isnot.friends`, `ls()`, and `rm()` to remove any object in the workspace that is not `friends`. 
```{r}
rm(list = ls()[isnot.friends])
```




## **Activity 4**

## Visualizing data with the ggplot2 package
Install and load the ggplot2 package
```{r}

```

Create a scatterplot of age by height with `ggplot()`. Refer to the ggplot2 cheat sheet (https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).
Hint: See `geom_point()`
```{r}

```

Create a line graph of age by height
```{r}

```

Create a boxplot of the variable `height`
```{r}

```

